Linear stability of magnetized massive protoplanetary disks

### 2014-04-07 paper/paper.tex
minor edits

\begin{abstract}
  Magneto-rotational instability (MRI) and gravitational instability
  (GI) are the two principle routes to turbulent angular momentum
  transport in accretion disks. Protoplanetary disks may develop 
  both. This paper aims to reinvigorate interest in the study of
  magnetized  massive protoplanetary disks, starting from the basic
  issue of stability. The local linear stability of a 
  self-gravitating, uniformly magnetized, differentially rotating, 
  three-dimensional disk subject to axisymmetric perturbations is
  calculated numerically. The formulation includes 
  resistivity. It is found that the reduction in the disk thickness
  by self-gravity (SG) can decrease MRI growth rates; the MRI
  becomes global in the vertical direction, and MRI modes 
  with small radial lengthscales are stabilized. In massive
  disks with layered resistivity, the MRI is not well-localized
  to regions where the Elsasser number exceeds unity. For MRI modes
  with radial lengthscales on the order of the disk thickness, SG
  can enhance density perturbations, an effect that becomes 
  significant in the presence of a strong toroidal field. In
  gravitationally unstable disks where GI and MRI growth rates are
  comparable, the character of unstable modes can transition smoothly
  between MRI and GI. Implications for non-linear simulations are
  discussed briefly.

>>>>

\begin{abstract}
  Magneto-rotational instability (MRI) and gravitational instability
  (GI) are the two principle routes to turbulent angular momentum
  transport in accretion disks. Protoplanetary disks may develop 
  both. This paper aims to reinvigorate interest in the study of
  magnetized  massive protoplanetary disks, starting from the basic
  issue of stability. The local linear stability of a 
  self-gravitating, uniformly magnetized, differentially rotating, 
  three-dimensional disk subject to axisymmetric perturbations is
  calculated numerically. The formulation includes 
  resistivity. It is found that the reduction in the disk thickness
  by self-gravity can decrease MRI growth rates; the MRI
  becomes global in the vertical direction, and MRI modes 
  with small radial lengthscales are stabilized. In massive
  disks with layered resistivity, the MRI is not well-localized
  to regions where the Elsasser number exceeds unity. For MRI modes
  with radial lengthscales on the order of the disk thickness, self-gravity 
  can enhance density perturbations, an effect that becomes 
  significant in the presence of a strong toroidal field. In
  gravitationally unstable disks where GI and MRI growth rates are
  comparable, the character of unstable modes can transition smoothly
  between MRI and GI. Implications for non-linear simulations are
  discussed briefly.

### 2014-04-07 paper/results_azi.tex
minor edits

\subsection{Ideal disks with MRI} 
We consider disk with $Q=0.2$ ($Q_\mathrm{2D}=~0.72$) and
$\beta=10$ in the limit of ideal MHD ($\Lambda_0=100$). Gravitational
instability is not expected because Fig. \ref{compare_growth3} shows
that even for $Q=0.18$, GI is surpressed for $\beta \lesssim 15$.

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 2cm 0cm
    0cm]{figures/compare_growth3_tilted_even.ps}  
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
    0.52cm]{figures/compare_growth3_tilted_odd.ps} 
  \caption{Growth rates in massive isothermal disks with $Q=~0.2$ and
    $\beta=10$ for a range of azimuthal field strengths $B_y/B_z$. The
    dots are solutions computed from the full problem, with the
    colorbar measuring the gravitational potential perturbation, while
    the solid curves are computed from the Cowling approximation. 
    For $B_y=0$, modes in top and bottom panels 
    $W^\prime(0)=0$ and $W(0)=0$, respectively.
    \label{compare_growth3_tilted}}
\end{figure}

>>>>

\subsection{Ideal disks with MRI} 
We consider disks with $Q=0.2$ ($Q_\mathrm{2D}=~0.72$) and
$\beta=10$ in the limit of ideal MHD ($\Lambda_0=100$). Gravitational
instability is not expected because Fig. \ref{compare_growth3} shows
that even for $Q=0.18$, GI is surpressed for $\beta \lesssim 15$.

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 2cm 0cm
    0cm]{figures/compare_growth3_tilted_even.ps}  
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
    0.52cm]{figures/compare_growth3_tilted_odd.ps} 
  \caption{MRI growth rates in isothermal disks with $Q=~0.2$ ($Q_\mathrm{2D}=0.72$) and 
    $\beta=10$ for a range of azimuthal field strengths $B_y/B_z$. The
    dots are solutions computed from the full problem, with the
    colorbar measuring the gravitational potential perturbation, while
    the solid curves are computed from the Cowling approximation. 
    For $B_y=0$, modes in top and bottom panels have
    $W^\prime(0)=0$ and $W(0)=0$, respectively.
    \label{compare_growth3_tilted}}
\end{figure}
